Anthropomorphism in Human-Robot Interaction: Nature, Dynamics and Underlying Cognitive Mechanisms

### 2014-11-28 paper.tex
Update section structure of part 4
Pass on the sections 3.2 -> 4

In this context, \emph{familiarization} means \emph{to get acquainted with}.
It is not to be confused with the \emph{familiarity thesis} that relates to the
projection of already known cognitive models onto the robot. We discuss this other
aspect in section~\ref{sec:cognition-neuroscience}.

The \emph{familiarization phase} starts at the peak of the \emph{novelty
effect}: anthropomorphic effects are at their maximum, the user thinks (s)he is
potentially facing an agent aiming at ``human-level intelligence'' (to take
McCarthy's words). This is a transient state that quickly vanishes, and the
projected anthropomorphism then starts to decrease while the human observes and
recognizes that the behavior of the robot is generally \emph{predictable} and
possibly \emph{deceptive} (for instance, it talks but does not understand when
we talk; it has eyes but does not recognize everyday objects we show; \etc).

\paragraph{Disruptive behaviors.}

We call \emph{disruptive behaviors} any behavior exhibited by the robot that
appears unexpected to the user: for instance, a robot may usually follow always
the same route to go from one place in a house to the other, but suddenly it
might change the route. The actual underlying reason may span from a bug to the
detection of a new obstacle, but as long as this reason is not immediately
intelligible to the user, the behavior counts as \emph{disruptive}. We further
discuss the role of disruptive behaviors in cognitive terms in section
\ref{sec:disruptive}.

@@

During the \textit{Stabilization Phase} the value of \Ant tends to stabilize over
a longer time, to reach a value that we call the \textit{Stabilized Level of
Anthropomorphism} (SLA), noted \SLAe. The value that the
SLA approaches is unique for each user interacting with a specific robot in a
given situation. \SLAe may approach $0$, when no anthropomorphic effects
are observed anymore, for instance when the user is interacting with the robot
in a routine way and he/she does not project any human-like traits unto the
robot, what \cite{takayama_toward_2011} calls the robot being
\textit{``invisible in use''}. \SLAe may as well remains above 0 if the user
continues to exhibit anthropomorphic behaviors when interacting with the robot
on the long run.

@@

@@

The model that we have introduced so far helps to represent and organize the
\emph{observations} of anthropomorphism. It does so at a conceptual level, and
with a special emphasis on the evolution over time.

This abstract model may contribute to a better understanding of the
\emph{phenomenology} of anthropomorphism, but it does not help to \emph{explain}
the phenomenon. This section proposes to look at the dynamics of
anthropomorphism in terms of \emph{cognitive correlates}: how human cognition
can explain the dynamic nature of our interactions with robots, and what are the
main cognitive layers upon which researchers can build to design better, deeper
human-robot interactions.

@@

\end{tikzpicture}
%}
\caption{\textbf{The three cognitive layers of anthropomorphism}: Layer 1 is the instinctive,
sub-cognitive identification of living peers. {\it Empathy} is characteristic
of this stage. After longer observation or short, non-contextualized interaction
(typically, a lab environment), the user enters Layer 2: the user projects a
mental model he/she is already familiar with onto the robot. After longer {\it
contextualized} interaction (typically, at home), the user enters Layer 3 of
anthropomorphism: the user composes a custom mental model of the robot,
based on experience. This leads to adapted interaction modalities, that may
still be anthropomorphic, or not. Note that the different layers coexist: Layer
1 and 2 still affect the human perception of the robot in a Layer 3 interaction.}
\label{fig:cognitivemodel}
\end{figure}

We propose three different cognitive layers (Figure~\ref{fig:cognitivemodel}),
which are not mutually exclusive but overlap over time. Note that these three
cognitive stages are related but do not exactly match the \emph{Initialization},
\emph{Familiarization} and \emph{Stabilization} phases introduced in our model
of the dynamics of anthropomorphism. We discuss this relation hereafter.

\subsubsection{Cognitive Processes and Layers\\}

@@

The intuitive anthropomorphization in the cognitive Layer 1 might be
characteristic of what \cite{takayama_perspectives_2012} calls the
\textit{in-the-moment} perspective on agency with a robot compared to the
\textit{reflective} perspective (see paragraph ``artifact-centric perspective''
in section \ref{sec:anthropomorphism}).  For instance, we reflectively might
\textit{not} perceive agency in a social robot, but it can feel quite
differently in the moment of interaction. \cite{takayama_perspectives_2012}
argues that neglecting to separate reflective perspectives from in-the-moment
perspectives of agency is one of the major sources of confusion when people talk
and write about anthropomorphism. There seems to be a disconnection between what
people consciously perceive and how they respond to stimuli that they may not
consciously perceive. This means that in the initial phase of interacting with a
novel device \textit{(initialization)} (see Figure \ref{fig:dynamics}) people
might not respond consciously but rather mindlessly \citep{nass_machines_2000},
and in turn anthropomorphize more. They start to respond in a more reflective
manner only during the \textit{familiarization} (in turn, the tendency to
anthropomorphize might decrease). This can be illustrated by the fact that
people tend to deny interacting with computational systems as if they were
people and yet they respond to computers in many ways that are remarkably
similar to how they respond to people \citep{reeves_media_1996}.
\cite{takayama_perspectives_2012} also applies a cognitive viewpoint on the two
different perspectives of in-the-moment \vs reflective to illustrate
differences. She states that in-the-moment perceptions of agency are largely
shaped by bottom-up perceptual processes, evoking very immediate responses (\eg
to the cues emitted by the artifact's design (``artifact-centric
perspective''). In contrast, according to Takayama, reflective perceptions are
more often shaped by top-down processes because of the nature of reflective
thought.

@@

The cognitive \emph{Layer 3} is gradually activated after a
\emph{contextualized} interaction.  A \emph{contextualized} interaction is
\emph{explicitly purposeful} (the purpose of the interaction, be it purely
entertainment, is explicit and conscious to the human), and takes place in an
environment that fosters a stronger cognitive (and possibly affective/social)
commitment from the human in the interaction (typically, at home). During this
interaction, the human iteratively restates and reshapes his/her behavioral and
mental model of the robot (\emph{How does the robot react to such and such
situation/input?  What does the robot know about me? About itself? About our
environment? What can the robot learn?}, etc.).

@@

\draw(0.5,5.5) node[anchor=south west] {\scriptsize \sc Init.};
\draw(2.5,5.5) node[anchor=south west] {\scriptsize \sc Familiarization};
\draw(7,5.5) node[anchor=south west] {\scriptsize \sc Stabilization};

@@

@@

\end{tikzpicture}
%}
\caption{\textbf{Towards a cognitive interpretation of the dynamics of
    anthropomorphism}: This diagram combines the figures~\ref{fig:dynamics}
    and~\ref{fig:cognitivemodel} to evidence their interplay: we suggest that
    the cognitive stages 1 and 2 (pre-cognitive anthropomorphism and projection
    of known existing cognitive models) first play a key role in the early
    perception of the robot, leading to high levels of anthropomorphic
    behaviors. Then, the user smoothly transitions from projecting already known
    models to building a new, specific cognitive model of the robot: this occurs
    while the user familiarizes with the robot and leads over time to a
    stabilized level of anthropomorphic behaviors. In the current state of
    robotic technology, this stabilization mostly reads as a \emph{decrease} of
    anthropomorphic perception, due to the user being able to explain and
    accurately predict the robot behavior after enough interaction time, hence
    perceiving the robot as a simple machine.}

@@

This mental process depends on the human understanding of the robot's inner
working, as well as his/her own tendency to anthropomorphize (the
\emph{personality} in IPA factor). At this stage, the \emph{perception} of the
robot (its shape for instance) and its original intended \emph{purpose} play
less of an important role. It is mostly a human-centric process.  The result of
this third phase is an iteratively adapted cognitive model of the robot.

@@

\paragraph{Relation to the dynamics of anthropomorphism\\}

@@

If the unexpected behavior is not planned, and perceived as such by the human
(case I, Table~\ref{fig:perceptionUnexpectedBehavior}), the human can interpret
that the robot is able to fail. If the robot explicitly states its failure (for
instance, by saying ``I'm lost!''), the behavior is then called
\emph{transparent}~\citep{kim_who_2006}, and suggests that the
robot has \emph{introspective} capabilities (it can reflect on its own internal
state), which may lead to higher anthropomorphic projections.  On the contrary,
if the robot shows no sign of recognizing its own failure, the user may ascribe
a lower level of anthropomorphism to the robot.

@@

In this paper, we have provided a broad account of the cognitive and
psychological viewpoints on anthropomorphism, with a novel focus on the
impact of time. Considering anthropomorphism as a social phenomenon that arises
in an interaction, we formulated the following three suggestions:

@@

\item \textbf{Anthropomorphism is based on multiple factors.} The three main
    factors that constitute anthropomorphism are the characteristics of the
    human, the characteristics of the robot, and the characteristics of the
    context of use. A temporal dimension is taken into account by the
    \textit{interaction history} $t$, which reflects the fact that
    anthropomorphism is also a dynamic phenomenon.

\item \textbf{Anthropomorphism is dynamic.} Looking at how experience and
    interaction evolve over time, we developed a qualitative model of the
    dynamics of anthropomorphism, which materializes a non-linear relationship
    between the observable \textit{anthropomorphic effects} \AntE of the human
    behavior and the \textit{interaction duration} $t$ (the actual time the user
    has spent interacting with the robot for real or imagined). This qualitative
    relationship starts at $t_{0}$ with the \textit{initial potential to
    anthropomorphize} (IPA), a function of the three factors previously
    discussed (point 1) that measures a pre-interaction form of
    anthropomorphism. On the other end, at $t_{\infty}$, we suggest that the
    interaction reaches a \textit{stabilized level of anthropomorphism} (SLA).
    How the curve evolves in between these two values is flexible and unique to
    each user, robot, and context. We however generally hypothesize an increase of
    \textit{anthropomorphic effects} driven by the \textit{novelty} of the
    robot, followed by a decrease of \AntE as the user \textit{familiarizes}
    him/herself with the robot and with using the robot.

\item \textbf{Anthropomorphism is multi-layered cognitive mechanism.} We have
    identified three successive and overlapping cognitive layers that underlie
    human-robot interaction and shape anthropomorphic behaviors.
    These layers (that are likely continuum rather than discrete levels) depend
    on one how typically human the robot is perceived, and how much the mental
    model that the user builds of the robot overlaps (or is inspired by) the
    mental model he/she has of another human. Each of the layers
    reflect different depths of cognitive engagement between the human and the
    robot.

@@

This approaches remain however fragile: in our experience, the weighting scheme
of the annotated behaviors mostly relies on a subjective assessment by the
experimenters, which makes it difficult to transfer to different experimental
settings. And it remains to be seen how self-reported assessment of
anthropomorphism (through questionnaires) can be meaningfully combined with
behavioral measurements, in a way that account for the dynamic nature of
anthropomorphism.

>>>>

The \emph{familiarization phase} starts at the peak of the \emph{novelty
effect}: anthropomorphic effects are at their maximum, the user thinks he/she is
potentially facing an agent aiming at ``human-level intelligence'' (to take
McCarthy's words). This is a transient state that quickly vanishes, and the
projected anthropomorphism then starts to decrease while the human observes and
recognizes that the behavior of the robot is generally \emph{predictable} and
possibly \emph{deceptive} (for instance, it talks but does not understand when
we talk; it has eyes but does not recognize everyday objects we show; \etc).

During this phase, we call \emph{disruptive behaviors} any behavior exhibited by
the robot that appears unexpected to the user: for instance, a robot usually
always follows the same route to go from one place in a house to the other, and
suddenly it changes the route. The actual underlying reason may span from a
bug to the detection of a new obstacle, but as long as this reason is not
immediately intelligible to the user, the behavior counts as \emph{disruptive}.
We further discuss the role of disruptive behaviors in cognitive terms in
section \ref{sec:disruptive}.

Note that, in this context, \emph{familiarization} means \emph{to get acquainted with}.
It is not to be confused with the \emph{familiarity thesis} that relates to the
projection of already known cognitive models onto the robot. We discuss this other
aspect in section~\ref{sec:cognition-neuroscience}.

@@

During the \textit{Stabilization Phase} the level of \AntE tends to stabilize over
a longer time, to reach a value that we call the \textit{Stabilized Level of
Anthropomorphism} (SLA), noted \SLAe. The value that the
SLA approaches is unique for each user interacting with a specific robot in a
given situation. \SLAe may approach $0$, when no anthropomorphic effects
are observed anymore, for instance when the user is interacting with the robot
in a routine way and he/she does not project any human-like traits unto the
robot, what \cite{takayama_toward_2011} calls the robot being
\textit{``invisible in use''}. \SLAe may as well remains above 0 if the user
continues to exhibit anthropomorphic behaviors when interacting with the robot
on the long run.

@@

The model that we have introduced so far helps to represent and organize the
\emph{observations} of anthropomorphism. It does so at a conceptual level, and
with a special emphasis on the evolution over time.  This abstract model may
contribute to a better understanding of the \emph{phenomenology} of
anthropomorphism, but it does not help to \emph{explain} the phenomenon. To this
end, we propose in the following section to explore the cognitive mechanisms at
play that account for anthropomorphism.

@@

This section proposes to look at the dynamics of anthropomorphism in terms of
their underlying cognitive mechanisms: how human cognition can explain the
dynamic nature of our interactions with robots, and what are the main cognitive
layers upon which researchers can build to design better, deeper human-robot
interactions.

@@

\end{tikzpicture}
%}
\caption{\textbf{The three cognitive layers of anthropomorphism}: Layer 1 is the
instinctive, sub-cognitive identification of living peers. {\it Empathy} is
characteristic of this stage. The Layer 2 materializes after longer
observation or short, non-contextualized interaction (typically, a lab
environment): the user projects a mental model he/she is already familiar
with onto the robot. After longer {\it contextualized} interaction
(typically, at home), the user enters Layer 3 of anthropomorphism: the user
composes a custom mental model of the robot, based on experience. This leads
to adapted interaction modalities, that may still be anthropomorphic, or
not. Note that the different layers coexist: Layer 1 and 2 continue to affect the
human perception of the robot in a Layer 3 interaction.}

\label{fig:cognitivemodel}
\end{figure}

We propose three different cognitive layers (Figure~\ref{fig:cognitivemodel}),
which are not mutually exclusive but overlap over time. Note that these three
cognitive stages are related but do not exactly match the \emph{Initialization},
\emph{Familiarization} and \emph{Stabilization} phases introduced in our model
of the dynamics of anthropomorphism. We discuss this relation in
section~\ref{sec:relations}.

@@

This intuitive form of anthropomorphization in the cognitive Layer 1 might be
characteristic of what \cite{takayama_perspectives_2012} calls the
\textit{in-the-moment} perspective on agency with a robot compared to the
\textit{reflective} perspective. For instance, we reflectively might
\textit{not} perceive agency in a social robot, but it can feel quite
differently in the moment of interaction. \cite{takayama_perspectives_2012}
argues that neglecting to separate reflective perspectives from in-the-moment
perspectives of agency is one of the major sources of confusion when people talk
and write about anthropomorphism. There seems to be a disconnection between what
people consciously perceive and how they respond to stimuli that they may not
consciously perceive. This means that in the initial phase of interacting with a
novel device (\textit{initialization} on Figure \ref{fig:dynamics}) people
might not respond consciously but rather mindlessly \citep{nass_machines_2000},
and in turn anthropomorphize more. They start to respond in a more reflective
manner only during the \textit{familiarization}, and in turn, the tendency to
anthropomorphize might decrease. This can be illustrated by the fact that
people tend to deny interacting with computers as if they were
people and yet they respond to them in many ways that are remarkably
similar to how they respond to people \citep{reeves_media_1996}.
\cite{takayama_perspectives_2012} also applies a cognitive viewpoint on the two
different perspectives of in-the-moment \vs reflective to illustrate
differences. She states that in-the-moment perceptions of agency are largely
shaped by bottom-up perceptual processes, evoking very immediate responses (\eg
to the cues emitted by the artifact's design -- the artifact-centric
perspective. In contrast, according to Takayama, reflective perceptions are
more often shaped by top-down processes because of the nature of reflective
thought.

@@

The cognitive \emph{Layer 3} is gradually activated after a
\emph{contextualized} interaction.  A \emph{contextualized} interaction is
\emph{explicitly purposeful} (the purpose of the interaction, be it purely
entertainment, is explicit and conscious to the human), and takes place in an
environment that fosters a stronger cognitive (and possibly affective/social)
commitment from the human in the interaction (typically, at home). During this
interaction, the human iteratively restates and reshapes his/her behavioral and
mental model of the robot (\emph{How does the robot react to such and such
situation/input?  What does the robot know about me? About itself? About our
environment? What can the robot learn?}, etc.).
This mental process depends on the human understanding of the robot's inner
working, as well as his/her own tendency to anthropomorphize (the
\emph{personality} in IPA factor). At this stage, the \emph{perception} of the
robot (its shape for instance) and its original intended \emph{purpose} play
less of an important role. It is mostly a human-centric process.  The result of
this third phase is a custom cognitive model of the robot.

@@

\draw(0.5,6.0) node[anchor=south west] {\scriptsize \sc Init.};
\draw(2.5,6.0) node[anchor=south west] {\scriptsize \sc Familiarization};
\draw(7,6.0) node[anchor=south west] {\scriptsize \sc Stabilization};

@@

\draw (-0.05, 2.1) -- (0.05, 2.1) node[anchor=east] {\IPAe};
\draw (1.8, 4.9) -- (1.8, 5.1) node[anchor=south] {\scriptsize \AntEMax};
\draw (10, 0.8) node[anchor=west] {\SLAe};

@@

\end{tikzpicture}
%}
\caption{\textbf{Towards a cognitive interpretation of the dynamics of
    anthropomorphism}: This diagram combines the figures~\ref{fig:dynamics}
    and~\ref{fig:cognitivemodel} to evidence their interplay: 
    the cognitive layers 1 and 2 (pre-cognitive anthropomorphism and projection
    of known existing cognitive models) play a key role in the early
    perception of the robot, leading to high levels of anthropomorphic
    behaviors. Then, the user smoothly transitions from projecting already known
    models to building a new, specific cognitive model of the robot: this occurs
    while the user familiarizes with the robot and leads over time to a
    stabilized level of anthropomorphic behaviors. In the current state of
    robotic technology, this stabilization mostly reads as a \emph{decrease} of
    anthropomorphic perception, due to the user being able to explain and
    accurately predict the robot behavior after enough interaction time, hence
    perceiving the robot as a simple machine.}

@@

@@

\subsection{Relation to the dynamics of anthropomorphism}
\label{sec:relations}

@@

If the unexpected behavior is not planned, and perceived as such by the human
(case I, Table~\ref{fig:perceptionUnexpectedBehavior}), the human can interpret
that the robot is able to fail. If the robot explicitly states its failure (for
instance, by saying ``I'm lost!''), the behavior is then called
\emph{transparent}~\citep{kim_who_2006}, and suggests that the robot has
\emph{introspective} capabilities (it can reflect on its own internal state),
which may lead to higher anthropomorphic projections.  On the contrary, if the
robot shows no sign of recognizing its own failure (a ``plain'' bug), the user
is likely to ascribe a lower level of anthropomorphism to the robot.

@@

This article provides a broad account of the cognitive and
psychological viewpoints on anthropomorphism, with a novel focus on the
impact of time. Considering anthropomorphism as a social phenomenon that arises
in an interaction, we formulated the following three suggestions:

@@

\item \textbf{Anthropomorphism is based on multiple factors.} The three main
    factors that constitute anthropomorphism are the characteristics of the
    human, the characteristics of the robot, and the characteristics of the
    context of use. Besides, a temporal dimension is taken into account by the
    \textit{interaction history} $t$, which reflects the fact that
    anthropomorphism is also a dynamic phenomenon.

\item \textbf{Anthropomorphism is dynamic.} Looking at how experience and
    interaction evolve over time, we developed a qualitative model of the
    dynamics of anthropomorphism, which materializes a non-linear relationship
    between the observable \textit{anthropomorphic effects} \AntE of the human
    behavior and the \textit{interaction duration} $t$ (the actual time the user
    has spent interacting with the robot for real or imagined). This qualitative
    relationship starts at $t_{0}$ with what we propose to call an \textit{initial potential to
    anthropomorphize} (IPA), a function of the three factors previously
    discussed (point 1) that measures a pre-interaction form of
    anthropomorphism. On the other end, at $t_{\infty}$, we suggest that the
    interaction reaches a \textit{stabilized level of anthropomorphism} (SLA).
    How the curve evolves in between these two values is flexible and unique to
    each user, robot, and context. We however generally hypothesize an increase of
    \textit{anthropomorphic effects} driven by the \textit{novelty} of the
    robot, followed by a decrease of \AntE as the user \textit{familiarizes}
    him/herself with the robot and with using the robot.

\item \textbf{Anthropomorphism is a multi-layered cognitive mechanism.} We 
    identify three successive and overlapping cognitive layers that underlie
    human-robot interaction and shape anthropomorphic behaviors.
    These layers (that are likely continuum rather than discrete levels) depend
    on ``how typically human'' the robot is perceived, and how much the mental
    model that the user builds of the robot overlaps (or is inspired by) the
    mental model he/she has of another human. Each of these layers
    reflect different depths of cognitive engagement between the human and the
    robot.

@@

This approache remains however fragile: in our experience, the weighting scheme
of the annotated behaviors mostly relies on a subjective assessment by the
experimenters, which makes it difficult to transfer to different experimental
settings. And it remains to be seen how self-reported assessment of
anthropomorphism (through questionnaires) can be meaningfully combined with
behavioral measurements, in a way that account for the dynamic nature of
anthropomorphism.
